Kindly Bent to Free Us

### 2019-11-14 kindly/lang.tex
Add description of the automatic region annotation.

>>>>

\subsection{Automatic region annotation}
\label{regionannot}

In the examples presented in \cref{motivation}, region annotations are optional.
In our formalization, regions must be syntactically explicit and annotated with
a nesting index and a scoped variable.
We now define a best-effort
syntactic code transformation $\RannotT{p}{p'}$ which
annotates programs automatically.
The input $p$ is a program with optional region annotations of the form $\region[]{}{e}$.
The output $p'$ is a program with explicit annotations
of the form $\region{\Sone x\BORROW}{e}$.
We give an informal presentation of our code transformation. The complete
definition is given in \cref{appendix:regionannot}. This code transformation
aims to find, for each borrow $\borrow{x}$, the biggest region satisfying
the following rules:
\begin{enumerate}
\item The region should contain at least $\borrow{x}$.
\item The region should never be larger than the scope of $x$.
\item An exclusive borrow $\borrow[\kaff]{x}$ should never share a region with any other borrow of $x$.
\item A use of $x$ can never be in the region of $\borrow{x}$.
\end{enumerate}
Following these rules, we start from the leafs and grow the region
associated to each borrow until it can't grow anymore, either due to a binding
or to conflicting uses.

Let us consider the following program with binders and function applications:

\[
  \setlength{\jot}{-1pt}
  \RannotT{
\begin{aligned}
  \lam{a}{}\ &
  \letin{x}{\app{f}{\borrow[\kun]{a}}}{}\\
  &g\ (\borrow[\kaff]{x});\\
  &f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})\\
\end{aligned}
}{
\begin{aligned}
  \lam{a}{}\region[1]{\Sone a\MBORROW}{\ &%
  \letin{x}{\app{f}{\borrow[\kaff]{a}}}{}\\
  &\region[2]{\Sone x\MBORROW}{g\ (\borrow[\kaff]{x})};\\
  &\region[2]{\Sone x\IBORROW}{f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})}\\
  &\!\!\!}
\end{aligned}
}
\]

Since the only use of $a$ is a borrow,
the region covers the entire lexical scope of $a$. For $x$ however, there are
multiple conflicting borrow and we must be more careful. We place a first
region around the exclusive borrow and its function application,
and a second region around both shared borrows. This placement of region
is optimal, as making any of the region bigger would cause a borrowing
error.

Naturally, these rules are best effort and it's sometimes necessary for the
programmer to annotate regions explicitly. We take such manual annotations
into account as an upper bound on all the contained regions.
In the following program, a manual annotation has been inserted to ensure
no borrow enters the reference $r$:

\[
  \setlength{\jot}{-1pt}
  \RannotT{
\begin{aligned}
  \letin{&r}{\operatorname{ref}\ 0}{}\\
  \lam{a}{}\ &
  set\ r\ \region[]{}{g\ (\borrow[\kun]{a})};\\
  &f\ (\borrow[\kun]{a}) 
\end{aligned}
}{
\begin{aligned}
  \letin{&r}{\operatorname{ref}\ 0}{}\\
  \lam{a}{}\ &
  set\ r\ \region[1]{\Sone a\IBORROW}{g\ (\borrow[\kun]{a})};\\
  &\region[1]{\Sone a\IBORROW}{f\ (\borrow[\kun]{a})}
\end{aligned}
}
\]

Following the rules, we could merge the two regions around the borrows
$\borrow[\kun]{a}$. However
the explicit annotation indicates that the region should stay around the closure
passed as argument of $set$. This is particularly useful to control captures
by imperative APIs.

This code transformation is purely syntactic and must be used before
typing. It only produces well-nested annotations: if $\region[n]{b}{\dots}$
is nested inside $\region[n']{b'}{\dots}$, then $n > n'$. Furthermore, there
is at most one region per borrow, and exactly one region per exclusive borrow.
In the rest of this article, we consider
that all terms have been annotated with this code transformation and respect
these properties.
